Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-16 paper.qmd
Improved some comments

When shapes are nonlinear in the NLDR layout, some edges could be long. It can also happen that distant centroids can be connected, particularly if clustering is present, which can result in long line segments. In order to generate a smooth surface in \gD{}, these long line segments should be removed when tuning the model fit. <!-- I don't think this is a thing anymore -->

@@

#### Measurement of capturing the data shape in 2-D

@@

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, UMAP was performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different (hyper-)~parameters. <!-- I think some of these panels aren't umap - if so saying "umap was performed on the first nine principal components" is not necessarily correct. I think its what the publication did but not what is in our figure --> You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

>>>>

@@

#### Measurement of capturing the data shape in \gD{}

@@

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different (hyper-)~parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.
